---
title: "Preparing Data with `exportRecordsTyped`"
subtitle: "and Frequently Asked Questions"
date: "`r Sys.Date()`"
toc: true
toc-title: "Contents"
output: pdf_document
---

# Introduction

The addition of `exportRecordsTyped` opened a great deal of flexibility and potential for customization when exporting data from REDCap and preparing them for analysis. The tasks of preparing data are broadly categorized into three phases

1. Missing Value Detection 
2. Field Validation
3. Casting Data

This document organizes discussion of these phases in the reverse order, reflecting the frequency of which questions are received regarding each phase. Throughout the document, examples are made using the Quality Assurance project used for testing `redcapAPI`.

```{r, result = 'hide'}
library(redcapAPI)
url <- "https://redcap.vanderbilt.edu/api/" # Our institutions REDCap instance

unlockREDCap(c(rcon = "TestRedcapAPI"), 
             envir = .GlobalEnv,
             keyring = "API_KEYs", 
             url = url)
```

\clearpage
# Casting Data

The default casting parameters were chosen with consideration for what is believed to be the most frequently desired results. The default casting list is shown in the appendix. It is inevitable that the circumstances of a particular project will necessitate customization. Furthermore, the decisions regarding default casting are inherently opinionated, and some users will prefer different castings. This section will discuss how to customize casting for field types as well as how to customize the casting of a single field. 

A full listing of the casting functions provided by `redcapAPI` are listed in the "Value" section of `?fieldValidationAndCasting`. 

## Customizing a Field Type Casting

Using the `cast` argument, the user may issue alternative casting instructions for any of the supported field types. In the following call, any fields having the type `date_` will be cast using the `as.Date()` function. Meanwhile, all other field types will be cast using the default casting list.

```{r, eval = FALSE}
exportRecordsTyped(rcon, 
                   cast = list(date_ = as.Date()))
```

Radio button and drop down fields are field type where users frequently need a value different than the default. In most cases, the user desires that these fields be cast to their coded values instead of the labeled values. Compare the results of these two commands:

```{r}
# Returns a factor with levels "Green", "Blue", and "Lavender"
Rec <- exportRecordsTyped(rcon, 
                          records = 1:3, 
                          fields = "dropdown_test")
Rec["dropdown_test"]

# Returns a factor with levels "1", "2", and "3"
Rec <- exportRecordsTyped(rcon, 
                          records = 1:3,
                          fields = "dropdown_test", 
                          cast = list(dropdown = castCode))
Rec["dropdown_test"]

# Returns a character value of the labeled values
Rec <- exportRecordsTyped(rcon, 
                          records = 1:3, 
                          fields = "dropdown_test", 
                          cast = list(dropdown = castLabelCharacter))
Rec["dropdown_test"]
```

It is also permissible to use self-made functions in casting. Consider the scenario where it is necessary to multiply a numeric field by 3 when performing the export. This may be accomplished by first defining a function then passing it to the override for the `number` field type.

Custom functions should have the arguments `x`, `field_name`, and `coding`. These arguments are necessary, even if they will not be used by the function. 

```{r}
multiply3 <- function(x, field_name, coding) as.numeric(x) * 3

# Return the actual values from the project
Rec <- exportRecordsTyped(rcon, 
                          records = 1:3, 
                          fields = "number_test")
Rec["number_test"]

# Return the values with the custom casting function
Rec <- exportRecordsTyped(rcon, 
                          records = 1:3, 
                          fields = "number_test", 
                          cast = list(number = multiply3))
Rec["number_test"]
```

It should be noted that applying a custom function in this way would impact all of the fields of type "number". It would be rare that such an outcome is desirable. These custom functions can also be written in a manner that impacts only one specific field. 

## Customizing a Casting for a Single Field

User-written functions used in casting overrides must contain the arguments `x`, `field_name`, and `coding`, even if these arguments are not intended to be used by the function. Their inclusion, however, makes it possible to write casting overrides that target only a specific field. In the previous section, a function was written that multiplied every field with the field type "number" by three during the export. By adding an `if` statement, a test can be performed against the field name and modifications can be applied only to the targeted field.

```{r}
multiply3_one_field <- function(x, field_name, coding){
  x <- as.numeric(x)
  if (field_name == "number_test") x * 3 # multiply target field by 3
  else x                                 # return other fields unaltered
}

Rec <- 
  exportRecordsTyped(rcon, 
                     records = 1:3, 
                     fields = c("number_test", "prereq_number"))
Rec[c("number_test", "prereq_number")]

Rec <- exportRecordsTyped(rcon, 
                          records = 1:3, 
                          cast = list(number = multiply3_one_field),
                          fields = c("number_test", "prereq_number"))
Rec[c("number_test", "prereq_number")]
```

Radio buttons and drop down fields are, again, field types where such customization is frequently needed. Consider the case of a drop down field where the coded values have special meaning, such as the numeric values of the coding represent numeric scores toward a composite quality of life index. However, other drop down fields in the project are desired to return the labeled values for categorical analysis. A user-defined function can be written to accommodate this scenario.

```{r}
special_cast_radio <- function(x, field_name, coding){
  if (field_name %in% "prereq_radio"){
    as.numeric(x)                    # Cast target field as numeric
  } else {
    castLabel(x, field_name, coding) # still uses the default for 
                                     # the non-targeted fields
  }
}

# Using the default casting
Rec <- exportRecordsTyped(rcon, 
                          records = 1:3, 
                          fields = c("radio_test", "prereq_radio"))
Rec[c("radio_test", "prereq_radio")]

# Use the user-defined function to change casting of one field
Rec <- exportRecordsTyped(rcon, 
                          records = 1:3, 
                          fields = c("radio_test", "prereq_radio"), 
                          cast = list(radio = special_cast_radio))
Rec[c("radio_test", "prereq_radio")]
```

\clearpage
# Field Validation

\clearpage
# Missing Value Detection

\clearpage
```{r, child = 'FAQ.Rmd'}
```

\clearpage
# Appendix

## Casting Field Types

* `calc`: Calculated fields.
* `checkbox`: Checkbox fields.
* `date_`: Text fields with the "Date" validation type. 
* `datetime_`: Text fields with the "Datetime" validation type.
* `datetime_seconds_`: Text fields with the "Datetime with seconds" validation type.
* `dropdown`: Drop down multiple choice fields.
* `float`: Text fields with the "Number" validation type.
* `form_complete`: Fields automatically added by REDCap indicating the completion status of the form.
* `int`: Text fields with the "Integer" validation type. This appears to be a legacy type, and integer appears to be used by more recent version of REDCap.
* `integer`: Text fields with the "Integer" validation type.
* `number`: Text fields with the "Number" validation type.
* `number_1dp`: Text fields with the "number (1 decimal place)" validation type.
* `number_1dp_comma_decimal`: Text fields with the "number (1 decimal place - comma as decimal)" validation type.
* `number_2dp`: Text fields with the "number (2 decimal place)" validation type.
* `number_2dp_comma_decimal`: Text fields with the "number (2 decimal place - comma as decimal)" validation type.
* `radio`: Radio button fields.
* `select`: Possible alias for `dropdown` or `radio`.
* `sql`: Fields that use a SQL query to make a drop down tools from another project.
* `system`: Fields automatically provided by REDCap for the project. These include `redcap_event_name`, `redcap_data_access_group`, `redcap_repeat_instrument`, and `redcap_repeat_instance`.
* `time_mm_ss`: Text fields with the "Time (MM:SS)" validation type.
* `time_hh_mm_ss`: Text fields with the "Time (HH:MM:SS)" validation type.
* `truefalse`: True - False fields.
* `yesno`: Yes - No fields.

## Default Casting List

```{r, eval = FALSE}
.default_cast <- list(
  date_                    = function(x, ...) as.POSIXct(x, format = "%Y-%m-%d"),
  datetime_                = function(x, ...) as.POSIXct(x, format = "%Y-%m-%d %H:%M"),
  datetime_seconds_        = function(x, ...) as.POSIXct(x, format = "%Y-%m-%d %H:%M:%S"),
  time_mm_ss               = function(x, ...) chron::times(ifelse(is.na(x),
                                                                  NA,
                                                                  paste0("00:",x)), 
                                                           format=c(times="h:m:s")),
  time_hh_mm_ss            = function(x, ...) chron::times(x, format=c(times="h:m:s")),
  time                     = function(x, ...) chron::times(gsub("(^\\d{2}:\\d{2}$)", 
                                                                "\\1:00", x), 
                                                           format=c(times="h:m:s")),
  float                    = as.numeric,
  number                   = as.numeric,
  number_1dp               = as.numeric, 
  number_1dp_comma_decimal = castDpNumeric(),
  number_2dp               = as.numeric, 
  number_2dp_comma_decimal = castDpNumeric(),
  calc                     = as.numeric,
  int                      = as.integer,
  integer                  = as.numeric,
  yesno                    = castLabel,
  truefalse                = function(x, ...) x=='1' | tolower(x) =='true',
  checkbox                 = castChecked,
  form_complete            = castLabel,
  select                   = castLabel,
  radio                    = castLabel,
  dropdown                 = castLabel,
  sql                      = NA, 
  system                   = castLabel
)
```