# Frequently Asked Questions

## How do I stop casting fields to factors?

_I used to be able to set `factors = FALSE` to prevent categorical values from being returned as factors. How do I do that with `exportRecordsTyped`?_

Users may substitute an alternate casting list specification within the call to `exportRecordsTyped`. `redcapAPI` provides two lists for this purpose: `default_cast_character` and `default_cast_no_factor`. These two lists are identical and may be used interchangeably.

```{r, eval = FALSE}
exportRecordsTyped(rcon, 
                   cast = default_cast_character)

exportRecordsTyped(rcon, 
                   cast = default_cast_no_factor)
```

Aside from not casting factors, all other settings in this list are identical to the default casting.

## How do I control the casting of `redcap_event_name`?

_In earlier versions of `redcapAPI`, the `redcap_event_name` field commonly returned the values such as `event_1_arm_1`, `event_2_arm_1`, etc. It now returns "fancy" values. How do I get the original behavior?_

The `redcap_event_name` field is one of the fields referred to as a "system" field. These fields are not part of the project's data dictionary, and are automatically returned by the API based on the configuration of the project. 

By default, `exportRecordsTyped` returns the "labeled" values of the event names. 

```{r}
exportRecordsTyped(rcon, 
                   fields = "redcap_event_name", 
                   records = 1:3)
```

This behavior can be changed using the `system` casting override (this will also affect the casting of other system fields). 

```{r}
exportRecordsTyped(rcon, 
                   fields = "redcap_event_name", 
                   records = 1:3, 
                   cast = list(system = castRaw))
```