---
title: "Using Offline Connections"
date: "`r Sys.Date()`"
output: pdf_document
---

# Introduction

The `redcapAPI` package works best when the user has permission to export and import data via the REDCap API. This requires the user to be added to a project, granted API permissions by the project owner, an API token requested, and the REDCap Administrator approving the API token request. There are many reasons why any one of these steps my not be completed which may leave the user without access to the API. 

"Offline connections" are a tool designed to provide the user with at least a subset the functionality available to API users. Most prominently, `exportRecordsTyped` has a method for offline connections that allows offline users to prepare data for analysis just as they would if an API connection were available. Instead of exporting the various data and meta data elements via the API, the offline connection is constructed using the comma separate value (CSV) files that may be downloaded from the REDCap user interface. 

# Basic Offline Connections

Offline connections are most commonly made using the CSV downloads from the REDCap project. In order to utilize `exportRecordsTyped`, the two essential downloads are the raw data and the data dictionary. For the best results, the data should be the raw, unlabeled data. Once these two files are downloaded, the offline connection is made by passing the file paths to the `offlineConnection` function. The `redcapAPI` package includes a set of files that may be used for demonstration. 

```{r}
suppressPackageStartupMessages(library(redcapAPI))

offline_file_dir <- system.file("extdata/offlineConnectionFiles", 
                                package = "redcapAPI")

records_file <- file.path(offline_file_dir, 
                          "TestRedcapApi_Records.csv")
metadata_file <- file.path(offline_file_dir, 
                           "TestRedcapAPI_DataDictionary.csv")

off_con <- offlineConnection(meta_data = metadata_file, 
                             records = records_file)

off_con
```

With the offline connection object established, the user is now able to prepare data for analysis as if using an API connection. 

```{r}
Records <- exportRecordsTyped(off_con)

# Show a subset of the formatted data
head(Records)[1:10]
```

# Getting Full Functionality from Offline Connections

With just the records and the data dictionary, the offline connection is able to provide the basic functionality of casting records. The casting performed so far, though, yields a warning that the REDCap URL was not provided to the connection object. 

As a consequence, links to the REDCap forms for data that failed validation cannot be constructed. In order to get more complete functionality from `exportRecordsTyped`, the user may also provide:

* The REDCap instance URL.
* The REDCap instance version number. 
* Events data.
* Project ID number.
* Whether the project is a classical or longitudinal project.

### URL and Version Number 

The URL passed to `offlineConnection` may be the same used by the API connection, even if the user does not have permissions to use the API.

The REDCap version number can be found at the bottom of the list of projects on the "My Projects" page. 

### Events Data

Events data can be downloaded from the REDCap user interface after opening the "My Projects" page, clicking on "Define My Events", and then selecting the "Dowload events (CSV)" option from the "Upload or Download" drop down.

Unfortunately, the events data downloaded from the user interface does not contain the event IDs, which are necessary if links to invalid data forms are desired. If the user wishes to have functional links, they will need to manually provide the event IDs. A data frame constructed in the following manner would be suitable to pass to `offlineConnection`.

```{r}
event_data <- 
  read.csv(file.path(offline_file_dir, 
                     "TestRedcapAPI_Events.csv"), 
           stringsAsFactors = FALSE)
event_data$event_id <- c(427837, 427838)
```

### Project ID and Longitudinal Status

The project ID number and longitudinal status are both values typically found in the project information exported via `exportProjectInformation`. Unfortunately, the user interface does not provide an equivalent download that can be used here. The user will have to provide these values manually to get full functionality from the offline connection. 

The project ID number is displayed in the user interface to the right of the project title. 

The user can determine if the status is longitudinal by looking at the Project Setup under the "Main Project Settings." If the "Use longitudinal data collection with defined events?" option is enabled, the project is longitudinal. 

The user need not provide the complete project information data frame to get most functionality from the offline connection. the following is enough to gain all functionality provided at the time of this writing.

```{r}
project_info <- data.frame(project_id = 167509, 
                           is_longitudinal = 1)
```

With all of these components in place, the offline connection can now be created with improved functionality. The warnings issued by this call indicate that not all of the fields permitted by the event data and the project information data have been provided. This is acceptable in this instance. `offlineConnection` tends to be pretty noisy about these types of issues; it has no way to know what the project configuration is and so chooses to draw the user's attention to potential issues quite liberally. 

```{r}
off_con <- offlineConnection(meta_data = metadata_file, 
                             records = records_file, 
                             url = "https://redcap.vanderbilt.edu/api/",
                             version = "13.10.3", 
                             events = event_data, 
                             project_info = project_info)

off_con
```

This time, when casting records for analysis, the warning about links to invalid data is absent, indicating that `redcapAPI` believes the links are likely to work. These are displayed in the report of invalid records.

```{r, results = 'asis'}
Records <- exportRecordsTyped(off_con)

reviewInvalidRecords(Records)
```

# Casting Records for Import

Offline connections may also be used to prepare data for import, even if the data wll not be imported using the API. For this type of casting, the offline connection only needs the data dictionary. To illustrate, the `Records` object previously prepared may not be prepared for import. After executing `castForImport`, the resulting data frame can be written to a CSV file and uploaded to the project via the REDCap user interface.

```{r}
off_con <- offlineConnection(meta_data = metadata_file)

ForImport <- castForImport(Records, 
                           rcon = off_con)
```

# Conclusion

Offline connections are an additional tool for managing REDCap data within R when the user does not have permission to use the API. The most anticipated use of offline connections is to give the user access to `exportRecordsTyped`. Without permission to use the API, the user should be prepared to manually look up project information in order to use the full power of `redcapAPI`. 